Fig2

### 2020-02-20 analysis/figure-2.Rmd
now with 1 fig showing total degs and a few examples

```{r setup, include=FALSE}
library(tidyverse)
library(cowplot)
source("../R/themes.R")

@@

```{r fig2, fig.width=7}

DEG_path <- "../results/DEseq2/"   # path to the data
DEG_files <- dir(DEG_path, pattern = "*DEGs") # get file names
DEG_pathfiles <- paste0(DEG_path, DEG_files)
DEG_files

@@

allDEG %>%
  group_by(sex, tissue, comparison) %>%
  summarize(n = n())

p1 <- allDEG %>%
  filter(comparison == "control_bldg") %>%
ggplot(aes(x = comparison,  fill = direction)) +
  geom_bar(position = "dodge") +
  facet_grid(tissue~sex) +
  theme_B3() +
  theme(#axis.text.x = element_text(angle = 45, hjust = 1),
         axis.text.x = element_blank(),
        legend.position = "none",
        strip.text.y = element_blank())  +
  guides(fill = guide_legend(nrow = 1)) +
  labs(x = " ", y = "Total Differentially Expressed Genes") +
  scale_fill_manual(values = colorscharnew,
                       name = " ",
                       drop = FALSE) +
  scale_color_manual(values = sexcolors)

p2 <- allDEG %>%
  filter(comparison != "control_bldg") %>%
ggplot(aes(x = comparison,  fill = direction)) +
  geom_bar(position = "dodge") +
  facet_grid(tissue~sex) +
  theme_B3() +
  theme(#axis.text.x = element_text(angle = 45, hjust = 1),
         axis.text.x = element_blank(),
        legend.position = "none",
        strip.placement = "right")  +
  guides(fill = guide_legend(nrow = 1)) +
  labs(x = "Sequential Parental Care Stages  ", y = NULL) +
  scale_fill_manual(values = colorscharnew,
                       name = " ", drop = FALSE) +
  scale_color_manual(values = sexcolors) +
  geom_text(stat='count', aes(label=..count..), vjust =-0.5, 
            position = position_dodge(width = 1),
            size = 2.5, color = "black")

forlegend <- allDEG %>%
ggplot(aes(x = comparison,  fill = direction)) +
  geom_bar(position = "dodge") +
   scale_fill_manual(values = colorscharnew,
                       name = "increased in", drop = FALSE) +
  scale_color_manual(values = sexcolors) +
  guides(color = F) +
  guides(fill= guide_legend(nrow =1)) +
  theme_B3() +
  theme(legend.position = "bottom")

mylegend <- get_legend(forlegend)

barplots <- plot_grid(p1,p2, nrow=1, rel_widths = c(0.9,3))

fig2 <- plot_grid(barplots, mylegend, nrow = 2, rel_heights = c(1,0.1))
fig2
```

>>>>

```{r setup, include=FALSE}
library(tidyverse)
library(cowplot)
library(kableExtra) 
library(ggtext) # for italic fonts
library(ggsignif) # fot stats on plots

source("../R/themes.R")

@@

```{r allDEGs}
DEG_path <- "../results/DEseq2/"   # path to the data
DEG_files <- dir(DEG_path, pattern = "*DEGs") # get file names
DEG_pathfiles <- paste0(DEG_path, DEG_files)
#DEG_files

@@

overlappingDEGs <- allDEG %>%
  filter(tissue == "hypothalamus",
         comparison != "control_bldg") %>%
  mutate(group = paste(sex, tissue,comparison, sep = ".")) %>%
  group_by(gene) %>%
  summarize(n = n()) %>%
  arrange(desc(n))
overlappingDEGs

candidategenes <- c("OXT", "AVP", "GNRH1",  "AR", "POMC", "AGRP",
                       "CRH", "AVPR1A", "AVPR1B", "AVPR2",
                       "CYP19A1", "DRD1", "DRD2", "PRL", "PRLR",
                    "SOX9")

nocontrol <- allDEG %>%
  filter(comparison != "control_bldg")

malegon <- nocontrol %>%
  filter(sex == "male", tissue == "gonad")
```

## variance stabilized gene expression  (vsd)

```{r vsd}
geneids <- read_csv("../metadata/00_geneinfo.csv")

vsd_path <- "../results/DEseq2/"   # path to the data
vsd_files <- dir(vsd_path, pattern = "*vsd.csv") # get file names
vsd_pathfiles <- paste0(vsd_path, vsd_files)
vsd_files

allvsd <- vsd_pathfiles %>%
  setNames(nm = .) %>% 
  map_df(~read_csv(.x), .id = "file_name")  %>% 
  dplyr::rename("gene" = "X1") %>% 
  pivot_longer(cols = L.G118_female_gonad_control:y98.o50.x_male_pituitary_inc.d3, 
               names_to = "samples", values_to = "counts")

getcandidatevsd <- function(whichgenes, whichtissue, whichsex){
  candidates  <- allvsd %>%
    filter(gene %in% whichgenes) %>%
    dplyr::mutate(sextissue = sapply(strsplit(file_name, '_vsd.csv'), "[", 1)) %>%
    dplyr::mutate(sextissue = sapply(strsplit(sextissue, '../results/DEseq2/'), "[", 2)) %>%
    dplyr::mutate(sex = sapply(strsplit(sextissue, '\\_'), "[", 1),
                tissue = sapply(strsplit(sextissue, '\\_'), "[", 2),
                treatment = sapply(strsplit(samples, '\\_'), "[", 4)) %>%
    dplyr::mutate(treatment = sapply(strsplit(treatment, '.NYNO'), "[", 1)) %>%
    dplyr::select(sex, tissue, treatment, gene, samples, counts) %>%
    filter(tissue == whichtissue, sex %in% whichsex)  %>%
    drop_na()
  candidates$treatment <- factor(candidates$treatment, levels = alllevels)
  return(candidates)
}

hypvsd <- getcandidatevsd(candidategenes, "hypothalamus", sexlevels)
pitvsd <- getcandidatevsd(candidategenes, "pituitary", sexlevels)
gonvsd <- getcandidatevsd(candidategenes, "gonad", sexlevels)
head(hypvsd)
head(pitvsd)
head(gonvsd)
```

```{r figurefunctions}

makebargraph <- function(whichtissue, myylab, lowlim){
  p <- allDEG %>%
    filter(tissue == whichtissue,
           comparison != "control_bldg") %>%
  ggplot(aes(x = comparison,  fill = direction)) +
    geom_bar(position = "dodge") +
    facet_grid(tissue~sex) +
    theme_B3() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none",
          strip.text.y = element_blank())  +
    guides(fill = guide_legend(nrow = 1)) +
    labs(x = "Sequential parental care stage comparisons", 
         y = myylab,
         subtitle = " ") +
  scale_fill_manual(values = allcolors,
                       name = " ",
                       drop = FALSE) +
  scale_color_manual(values = allcolors)  #+
 # geom_text(stat='count', aes(label=..count..), vjust =-0.5, 
 #           position = position_dodge(width = 1),
 #           size = 2.5, color = "black")  +
 # ylim(lowlim, higherlim)
  return(p)
}

makeboxplots <- function(df, whichgene, myylab, whichsex){
  p <- df %>%
    filter(treatment %in% charlevels,
           gene %in% whichgene,
           sex %in% whichsex) %>%
    filter(treatment != "control") %>%
    ggplot(aes(x = treatment, y = counts, fill = treatment, color = sex)) +
    geom_boxplot() + 
    #geom_point() +
    facet_wrap(~sex) +
    theme_B3() +
    scale_fill_manual(values = allcolors) +
    scale_color_manual(values = allcolors) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "none") +
    labs(y = myylab , x = "Parental stage", subtitle = "") +
    theme(axis.title.y = element_markdown())
  return(p)
}
```

## HPG

```{r fig2, fig.width=7, fig.height=7}

allDEG %>%
  filter(gene %in% candidategenes, 
         tissue == "hypothalamus") %>%
  group_by(sex, tissue, comparison) %>%
  summarize(genes = str_c(gene, collapse = " ")) %>%
  pivot_wider(names_from = comparison, values_from = genes ) %>%
  select(sex, tissue, control_bldg,  
        inc.d9_inc.d17, hatch_n5)  %>%
  arrange(sex, tissue)

write.csv(table1, "../results/table-1.csv", row.names = F)

# hyp
p1 <- makebargraph("hypothalamus","Hypothalamic DEGs") + theme(axis.text.x = element_blank(), axis.title.x = element_blank())
p2 <- makeboxplots(hypvsd, "AR","*AR* expression", "male") +
              geom_signif(comparisons=list(c("inc.d9", "inc.d17")), annotations= "*", 
              y_position = 8, tip_length = 0.005, vjust = 0.5, color= "black" , textsize = 3 ) +
              theme(axis.text.x = element_blank(), axis.title.x = element_blank())

p3 <- makeboxplots(hypvsd, "DRD1","*DRD1* expression", "female") +
              geom_signif(comparisons=list(c("hatch", "n5")), annotations= "*", 
              y_position = 10, tip_length = 0.005, vjust = 0.5, color= "black" , textsize = 3 ) +
              #geom_signif(comparisons=list(c("control", "bldg")), annotations= "*", 
              #y_position = 10, tip_length = 0.005, vjust = 0.5, color= "black" , textsize = 3 ) + 
              theme(axis.text.x = element_blank(), axis.title.x = element_blank())

p23 <- plot_grid(p3,p2, labels = c("b", "c"), label_size = 12)

# pit

p4 <- makebargraph("pituitary","Pituitary DEGs")  +  theme(axis.text.x = element_blank(), axis.title.x = element_blank(), strip.text = element_blank())
p5 <- makeboxplots(pitvsd, "PRL","*PRL* expression", "male") +  theme(axis.text.x = element_blank(), axis.title.x = element_blank(), strip.text = element_blank())+
              #geom_signif(comparisons=list(c("control", "bldg")), annotations= "*", 
              #y_position = 21.5, tip_length = 0.005, vjust = 0.5, color= "black" , textsize = 3 ) +
              geom_signif(comparisons=list(c("inc.d9", "inc.d17")), annotations= "*", 
              y_position = 21.5, tip_length = 0.005, vjust = 0.5, color= "black" , textsize = 3 ) 
p6 <- makeboxplots(pitvsd, "PRL","*PRL* expression", "female")  +  theme(axis.text.x = element_blank(), axis.title.x = element_blank(), strip.text = element_blank()) +
              geom_signif(comparisons=list(c("inc.d9", "inc.d17")), annotations= "*", 
              y_position = 21, tip_length = 0.005, vjust = 0.5, color= "black" , textsize = 3 ) +
              #geom_signif(comparisons=list(c("control", "bldg")), annotations= "*", 
              #y_position = 21, tip_length = 0.005, vjust = 0.5, color= "black" , textsize = 3 ) +
              geom_signif(comparisons=list(c("hatch", "n5")), annotations= "*", 
              y_position = 21, tip_length = 0.005, vjust = 0.5, color= "black" , textsize = 3 )

p56 <- plot_grid(p6,p5, labels = c("e", "f"), label_size = 12)

# gon

p7 <- makebargraph("gonad","Gonadal DEGs") +  theme(strip.text = element_blank())
p8 <- makeboxplots(gonvsd, "SOX9","*SOX9* expression", "male")  +  theme(strip.text = element_blank()) +
              geom_signif(comparisons=list(c("lay", "inc.d3")), annotations= "*", 
              y_position = 7.5, tip_length = 0.005, vjust = 0.5, color= "black" , textsize = 3 )
p9 <- makeboxplots(gonvsd, "AVPR1A","*AVPR1A* expression", "female") +  theme(strip.text = element_blank()) +
              #geom_signif(comparisons=list(c("control", "bldg")), annotations= "*", 
              #y_position = 9.5, tip_length = 0.005, vjust = 0.5, color= "black" , textsize = 3 ) +
              geom_signif(comparisons=list(c("lay", "inc.d3")), annotations= "*", 
              y_position = 9.5, tip_length = 0.005, vjust = 0.5, color= "black" , textsize = 3 ) +
              geom_signif(comparisons=list(c("inc.d3", "inc.d9")), annotations= "*", 
              y_position = 9.5, tip_length = 0.005, vjust = 0.5, color= "black" , textsize = 3 )

p89 <- plot_grid(p9,p8, labels = c("h", "i"), label_size = 12)

a <- plot_grid(p1, p23, nrow = 1, rel_heights = c(0.5,0.5), labels = c("a", " "), label_size = 12)
b <- plot_grid(p4, p56, nrow = 1, rel_heights = c(0.5,0.5), labels = c("d", " "), label_size = 12)
c <- plot_grid(p7, p89, nrow = 1, rel_heights = c(0.5,0.5), labels = c("g", " "), label_size = 12)
plot_grid(a,b,c, nrow = 3, rel_heights = c(1,0.9,1.3))

```

```{r}
allDEG %>%
  group_by(tissue, comparison) %>%
  summarize(totalDEGs = n()) %>%
  arrange(tissue, comparison)
```

```{r writefiles}
table1 <- allDEG %>%
  filter(gene %in% candidategenes) %>%
  group_by(sex, tissue, comparison) %>%
  summarize(genes = str_c(gene, collapse = " ")) %>%
  pivot_wider(names_from = comparison, values_from = genes ) %>%
  select(sex, tissue, control_bldg, lay_inc.d3, inc.d3_inc.d9,
        inc.d9_inc.d17, hatch_n5)  %>%
  arrange( tissue, sex)
table1

write_csv(table1, "../results/table1.csv")
```
